\documentclass{article}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{natbib}
\usepackage{color}
\bibliographystyle{abbrvnat}

\newtheorem{lemma}{Lemma}

% These macros are borrowed from TAOCPMAC.tex
\newcommand{\slug}{\hbox{\kern1.5pt\vrule width2.5pt height6pt depth1.5pt\kern1.5pt}}
\def\xskip{\hskip 7pt plus 3pt minus 4pt}
\newdimen\algindent
\newif\ifitempar \itempartrue % normally true unless briefly set false
\def\algindentset#1{\setbox0\hbox{{\bf #1.\kern.25em}}\algindent=\wd0\relax}
\def\algbegin #1 #2{\algindentset{#21}\alg #1 #2} % when steps all have 1 digit
\def\aalgbegin #1 #2{\algindentset{#211}\alg #1 #2} % when 10 or more steps
\def\alg#1(#2). {\medbreak % Usage: \algbegin Algorithm A (algname). This...
  \noindent{\bf#1}({\it#2\/}).\xskip\ignorespaces}
\def\kalgstep#1.{\ifitempar\smallskip\noindent\else\itempartrue
   \hskip-\parindent\fi
   \hbox to\algindent{\bf\hfil #1.\kern.25em}%
   \hangindent=\algindent\hangafter=1\ignorespaces}

\newcommand{\algstep}[3]{\kalgstep #1 [#2] #3 }
\newenvironment{taocpalg}[3]{%
\vspace{1em}%
\algbegin Algorithm #1. ({#2}). #3 }
{\vspace{1em}}


\DeclareMathOperator*{\npOnes}{Ones}
\DeclareMathOperator*{\npZeros}{Zeros}
\DeclareMathOperator*{\argmax}{argmax}

\begin{document}

\begin{abstract}
The Li and Stephens model is one of the key techiques underpinning modern
genomics. The ability to solve the model exactly for large number of
sites has made it very attractive. But it scales linearly in the number
of samples, which is now far too slow. People are using approximations
to address this. We introduce an algorithm that uses the succinct
tree sequence structure that solves the Li and Stephens model exactly
in logarithmic time (under certain assumptions about the structure
of the trees). We show how this algorithm out-performs existing implementations
and enables applications in the soon-to-be available megasample data sets.

\end{abstract}


\section{Introduction}
Hidden Markov Models (HMMs) are one of the key techniques in classical
bioinformatics~\citep{durbin1998biological}.

\begin{itemize}
\item Very large datasets.
\item Li and Stephens~\citep{li2003modeling,donnelly2010coalescent} is important. Cite Donnely and
Leslie. Lots of applications.
\item Much recent interest. Method improving performance with BWT~\citep{lunter2018haplotype}
and also \citep{rosen2018average}.
\item Tree sequences.
\citep{kelleher2016efficient,kelleher2018efficient,kelleher2018inferring,haller2018tree}.
\item Paper roadmap.
\end{itemize}

\section{Li and Stephens model}

\begin{itemize}
\item Description of the LS model with diagram based on the variant matrix.
\item Viterbi algorithm listing using variant matrix.
\item Quick algorithm analysis.
\item Possibly forward/backward also?
\end{itemize}

The transition probability is given by
\begin{equation}\label{eq-recomb-proba}
Q(j, k, \ell) = \begin{cases}
1 - r_\ell + r_\ell / n & \text{if $j = k$} \\
r_\ell / n & \text{otherwise}
\end{cases}
\end{equation}
where $Q(j, k, \ell)$ is the probability of transition from haplotype $j$
to haplotype $k$ at locus $\ell$, and $r_\ell$ is the recombination `weight'
between site $\ell$ and $\ell - 1$. [Is this the right way around? Work out
the corner case at 0/m.]


[Note: there are different ways in which we can calibrate the mutation and
recombination weights. Need to discuss this and put some bounds on what
the weights really mean and how they interplay between mutation and
recombination.]

The emission probability is given by
\begin{equation}\label{eq-mut-proba}
E(j, \ell) = \begin{cases}
1 - \mu_\ell / (a_\ell - 1)  & \text{if $H_{j, \ell} = h_{\ell}$} \\
\mu_\ell & \text{otherwise}
\end{cases}
\end{equation}
where $\mu_\ell$ is the probability of mutatating from one of the
$a_\ell$ alleles to a different allele at site $\ell$.


\begin{taocpalg}{M}{Matrix based Li and Stephens Viterbi algorithm.
Given a haplotype of $h$ of length $m$ and a reference panel $H$
of $n$ haplotypes, return a path $P$ with maximum probability given
the recombination and mutation weights, $\rho$ and $\mu$}

\algstep{M1.}{Initialisation.}{
Set $V \leftarrow \npOnes(n)$, $I \leftarrow \npZeros(m)$ and
$P\leftarrow \npZeros(m)$.
Then set $T_\ell \leftarrow \{\}$ for
$0 \leq \ell < m$ and finally set $\ell \leftarrow 0$. }

\algstep{M2.}{Normalise.}{
Set $I_\ell \leftarrow \argmax(V)$, $V \leftarrow V / V_{I_\ell}$,
$p_{\neq} \leftarrow \rho_\ell / n$ and $j \leftarrow 0$.}

\algstep{M3.}{Transition.}{
Set $p_t \leftarrow  (1 - \rho_\ell + \rho_\ell / n)V_j$.
Then, if $p_{\neq} > p_t $ set $p_t \leftarrow p_{\neq}$
and $T_\ell \leftarrow T_\ell \cup \{j\}$.
}

\algstep{M4.}{Emission.}{
If $H_{j, \ell} = h_\ell$ set $p_e \leftarrow 1 - (a_\ell - 1)\mu_\ell$;
otherwise, set $p_e \leftarrow \mu_\ell$.
}

\algstep{M5.}{Haplotype loop.}{
Set $V_{j} \leftarrow p_{t} p_{e}$ and $j \leftarrow j + 1$.
If $j < n$ go to step M3.
}

\algstep{M6.}{Site loop.}{
Set $\ell \leftarrow \ell + 1$.
If $\ell < m$ go to step M2.
}

\algstep{M7.}{Initialise Traceback.}{
Set  $\ell \leftarrow m - 1$ and $P_\ell \leftarrow
\argmax(V)$. }

\algstep{M8.}{Traceback.}{
Set $j \leftarrow P_\ell$. If $j \in T_\ell$, set $j \leftarrow I_\ell$.
Then set $P_{l - 1} \leftarrow j$ and $\ell \leftarrow \ell - 1$.
If $l > 0$, go to M8.
}

\end{taocpalg}

\textcolor{red}{TODO change the description here to talk about the
traceback set. currently out of date.}

In step M1 we set up main algorithm state. First we set up
the Viterbi vector $V$ where $V_j$ is the scaled likelihood of matching
with haplotype $j$ at the previous site. We also setup the traceback matrix
$T$, where $T_{j, \ell}$ encodes the most likely haplotype to copy from at site
$\ell - 1$ assuming that we are copying from haplotype $j$ at site $\ell$.
Finally, we set the site index variable $\ell$ to start iterating from the
left of the sequence.

Step M2 first normalises the Viterbi vector $V$ by finding the index $i$
of the most likely haplotype and then dividing all other values by this. We
then create the Viterbi vector for the current site $V'$, and begin
iterating over the haplotypes by setting $j$ to zero. Step M3 computes
the probabilyt of recombination, and is the heart of the LS algorithm.
We begin by setting $x$ to likelihood of copying from the current
haplotype without recombination, which is given by the Viterbi value
for haplotype $j$ from the previous site multiplied by the
probability of no recombination in \eqref{eq-recomb-proba}. We also
compute the probability of recombining away from haplotype $j$
to the best haplotype at the previous site, $i$. As we have already
rescaled $V$ such that $V_i = 1$, we know that this probability is
simply the probability of recombination \eqref{eq-recomb-proba}. Then,
if the probability of not recombining is larger than the probability
of recombining, we set the transition probability $p_t$ to $x$
and update the traceback matrix to note that we should stay on haploytype
$j$ at the current site (should be already be copying from haplotype $j$).
Otherwise, if the probability of recombination is greater, then we set the
transition probability accordingly and update $T_{j, \ell}$ so that we
transition to best haplotype $i$ at the previous site if we are already
copying from haplotype $j$.

Step M4 then computes the emission probability for the haplotype $j$
at the current site. [EXPLAIN] Steps M5 and M6 then complete the loops
for sites and haplotypes, respectively, after updating the Viterbi
vectors $V$ and $V'$.

When we get to step M7 we have completed the forward iteration over
all $m$ sites, and considered each of the $n$ haplotypes at each site.
The Viterbi vector $V$ contains the probabality of each of the haplotypes
of matching $h$ at site $m - 1$, conditioned on all of the previous sites.
We can then use the standard HMM approach of ``tracing back'' through
all of the sites to generate the best path $P$ for $h$ through the
reference panel $H$.

\textcolor{red}{Add diagram showing an example. We want to show the
traceback sets, and ideally the Viterbi matrices. We want to see that
there's actually a limited number of distinct values in there.
It's not true in general that we have only 3 distinct values---this arises
in the special case of equal recombination rates and with rare mutation}

\section{Tree sequence algorithm}

\begin{itemize}
\item Brief description of the tree sequence basics. Nodes, edges, etc.
\item Viterbi algorithm listing using tree sequences.
\item Algorithm analysis.
\end{itemize}

\section{Performance}

\begin{itemize}
\item Comparison of LS algorithm implementations.
\end{itemize}


\section{Discussion}

\bibliography{references}

\end{document}
